Why aren't significance tests commonly used for linear regression diagnostics?

### 2022-10-04 paper_comparison/paper_comparison/paper_comparison.Rmd
merged conflicts

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  echo = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%")
```

@@

Compared to non-linearity and heteroskedasticity, non-normality is usually harder to detect from a residual plot since scatter plot do not readily reveal the marginal distribution. A favourable graphical summary for this task is the quantile-quantile plot. However, for a consistent comparison, residual plot will be the focus of this paper. Besides, it is important to note that not all regression models assume normality for the error term, but a certain amount do including the classical normal linear regression model. In the case that the normality assumption is violated, it is expected to observe data points do not center around the horizontal axis and there is an uneven distribution of the number points at both below and above the horizontal axis. For example, given a skewed error distribution, fewer data points and more outliers are on one side of the horizontal axis as shown at the bottom-right of Figure \ref{fig:commondepartures}.

@@

Example residual plots given in Figure \ref{fig:commondepartures} are examined by the corresponding model structural test, Breusch-Pagan test and Shapiro--Wilk test as shown in Table \ref{tab:example-residual-plot-table}. In the example, both the Breusch-Pagan test and the Shapiro--Wilk test rejects the null hypothesis for departures that they do not intend to examine. As discussed in @cook1982residuals, most residual-based tests for a particular type of departure from model assumptions are sensitive to other types of departures. It is likely the null hypothesis is correctly rejected but for the wrong reason, which is known as the "Type III error". Additionally, outliers will often incorrectly trigger the rejection of the null hypothesis despite when majority of the residuals are well-behaved [@cook_applied_1999]. This can be largely avoided in diagnostic plots as experienced analysts can evaluate the acceptability of assumptions flexibly, even in the presence of outliers.

@@

## Experiment overview

@@

Experiment I is designed to study the ability of human subjects to detect the effect of a non-linear term $\boldsymbol{z}$ which is a probabilist's Hermite polynomial [Herimite ref here] of another random vector $\boldsymbol{x}$ in a two variable statistical model formulated as:

@@

```{r different-shape-of-herimite, fig.cap="Example residual plots of the linear regression model used in experiment I. Four different values of $j$ are included in the experiment results in four different shapes of residuals."}
stand_dist <- function(x) (x - min(x))/max(x - min(x)) * 2 - 1

VI_MODEL$plot(ori_dat) + 
  ggtitle(quote(He[2]:"U shape")) + 
  theme_light() -> p1

VI_MODEL$plot(poly_model(shape = 2, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, sigma = 0.05)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote(He[3]:"S shape")) + 
  theme_light() -> p2

VI_MODEL$plot(poly_model(shape = 3, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, sigma = 0.05)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote(He[6]:"M shape")) + 
  theme_light() -> p3

VI_MODEL$plot(poly_model(shape = 4, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, sigma = 0.05)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote(He[18]:"Triple-U shape")) + 
  theme_light() -> p4

@@

```{r different-dist, fig.cap="Example residual plots of the linear regression model used in experiment I. Four different distribution of $x_{raw}$ are used in the experiment to provide various visual patterns."}
ori_dat <- poly_model(shape = 1, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, sigma = 0.5)$gen(300)

VI_MODEL$plot(ori_dat) + 
  ggtitle(quote(U(-1,1))) + 
  theme_light() -> p1

VI_MODEL$plot(poly_model(shape = 1, x = {raw_x <- rand_normal(sigma = 0.3); closed_form(~stand_dist(raw_x))}, sigma = 0.5)$gen(300, computed = select(ori_dat, e))) +
  ggtitle(quote(N(0,0.3^2))) + 
  theme_light() -> p2

VI_MODEL$plot(poly_model(shape = 1, x = {raw_x <- rand_lognormal(sigma = 0.6); closed_form(~stand_dist(raw_x/3 - 1))}, sigma = 0.5)$gen(300, computed = select(ori_dat, e))) +
  ggtitle(quote(lognormal(0,0.6^2)/3)) + 
  theme_light() -> p3

VI_MODEL$plot(poly_model(shape = 1, x = rand_uniform_d(k = 5, even = TRUE), sigma = 0.5)$gen(300, computed = select(ori_dat, e))) +
  ggtitle(quote(U~"{1,5}")) + 
  theme_light() -> p4

@@

```{r different-shape-of-heter, fig.cap="Example residual plots of the linear regression model used in experiment II. Three different shapes (a = -1, 0, 1) are used in the experiment to create left-triangle shape, butterfly shape and right-triangle shape."}
stand_dist <- function(x) (x - min(x))/max(x - min(x)) * 2 - 1

@@

VI_MODEL$plot(ori_dat) +
  ggtitle(quote("a = -1: Left-triangle shape")) +
  theme_light() -> p1

VI_MODEL$plot(heter_model(a = 0, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, b = 128)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote("a = 0: Butterfly shape")) +
  theme_light() -> p2

VI_MODEL$plot(heter_model(a = 1, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, b = 128)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote("a = 1: Right-triangle shape")) +
  theme_light() -> p3

patchwork::wrap_plots(p1, p2, p3, ncol = 2)
```

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE, 
  warning = FALSE, 
  echo = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  fig.align = "center")
```

@@

Compared to non-linearity and heteroskedasticity, non-normality is usually harder to detect from a residual plot since scatter plot do not readily reveal the marginal distribution. A favourable graphical summary for this task is the quantile-quantile plot. However, for a consistent comparison, residual plot will be the focus of this paper. Besides, it is important to note that not all regression models assume normality for the error term, but a certain amount do including the classical normal linear regression model. In the case that the normality assumption is violated, it is expected to observe data points do not center around the horizontal axis and there is an uneven distribution of the number points at both below and above the horizontal axis. For example, given a skewed error distribution, fewer data points and more outliers are on one side of the horizontal axis as shown in plot D of Figure \ref{fig:residual-plot-common-departures}.

@@

Example residual plots given in Figure \ref{fig:residual-plot-common-departures} are examined by the corresponding model structural test, Breusch-Pagan test and Shapiro--Wilk test as shown in Table \ref{tab:example-residual-plot-table}. In the example, both the Breusch-Pagan test and the Shapiro--Wilk test rejects the null hypothesis for departures that they do not intend to examine. As discussed in @cook1982residuals, most residual-based tests for a particular type of departure from model assumptions are sensitive to other types of departures. It is likely the null hypothesis is correctly rejected but for the wrong reason, which is known as the "Type III error". Additionally, outliers will often incorrectly trigger the rejection of the null hypothesis despite when majority of the residuals are well-behaved [@cook_applied_1999]. This can be largely avoided in diagnostic plots as experienced analysts can evaluate the acceptability of assumptions flexibly, even in the presence of outliers.

@@

<!-- ## Experiment overview-->

@@

```{r different-shape-of-herimite, fig.height = 2, fig.cap="Polynomial forms generated for the residual plots used in experiment I. The four shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$."}
stand_dist <- function(x) (x - min(x))/max(x - min(x)) * 2 - 1

VI_MODEL$plot(ori_dat) + 
  ggtitle(quote(He[2]:"U")) + 
  theme_light() -> p1

VI_MODEL$plot(poly_model(shape = 2, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, sigma = 0.05)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote(He[3]:"S")) + 
  theme_light() -> p2

VI_MODEL$plot(poly_model(shape = 3, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, sigma = 0.05)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote(He[6]:"M")) + 
  theme_light() -> p3

VI_MODEL$plot(poly_model(shape = 4, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, sigma = 0.05)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote(He[18]:"Triple-U")) + 
  theme_light() -> p4

patchwork::wrap_plots(p1, p2, p3, p4, ncol = 4)
```

Experiment I is designed to study the ability of human subjects to detect the effect of a non-linear term $\boldsymbol{z}$ constructed using Hermite polynomials [Hermite ref here] on random vector $\boldsymbol{x}$ formulated as:

@@

@@

```{r different-dist, fig.height = 2, fig.cap="Variations in fitted values, that might affect perception of residual plots. Four different distribution of $x_{raw}$ are used in the experiment to provide various visual patterns."}
ori_dat <- poly_model(shape = 1, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, sigma = 0.5)$gen(300)

VI_MODEL$plot(ori_dat) + 
  ggtitle("A. Uniform") + #quote(U(-1,1))) + 
  theme_light() -> p1

VI_MODEL$plot(poly_model(shape = 1, x = {raw_x <- rand_normal(sigma = 0.3); closed_form(~stand_dist(raw_x))}, sigma = 0.5)$gen(300, computed = select(ori_dat, e))) +
  ggtitle("B. Normal") + #quote(N(0,0.3^2))) + 
  theme_light() -> p2

VI_MODEL$plot(poly_model(shape = 1, x = {raw_x <- rand_lognormal(sigma = 0.6); closed_form(~stand_dist(raw_x/3 - 1))}, sigma = 0.5)$gen(300, computed = select(ori_dat, e))) +
  ggtitle("C. Skewed") + #quote(lognormal(0,0.6^2)/3)) + 
  theme_light() -> p3

VI_MODEL$plot(poly_model(shape = 1, x = rand_uniform_d(k = 5, even = TRUE), sigma = 0.5)$gen(300, computed = select(ori_dat, e))) +
  ggtitle("D. Discrete") + #quote(U~"{1,5}")) + 
  theme_light() -> p4

patchwork::wrap_plots(p1, p2, p3, p4, ncol = 4)
```

@@

```{r different-shape-of-heter, fig.height=3, fig.cap="Heteroskedasticity forms used in experiment II. Three different shapes (a = -1, 0, 1) are used in the experiment to create left-triangle, butterfly and right-triangle shapes."}
stand_dist <- function(x) (x - min(x))/max(x - min(x)) * 2 - 1

@@

VI_MODEL$plot(ori_dat) +
  ggtitle(quote("A. a = -1: Left-triangle")) +
  theme_light() -> p1

VI_MODEL$plot(heter_model(a = 0, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, b = 128)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote("B. a = 0: Butterfly")) +
  theme_light() -> p2

VI_MODEL$plot(heter_model(a = 1, x = {raw_x <- rand_uniform(-1, 1); closed_form(~stand_dist(raw_x))}, b = 128)$gen(300, computed = select(ori_dat, x, e))) +
  ggtitle(quote("C. a = 1: Right-triangle")) +
  theme_light() -> p3

patchwork::wrap_plots(p1, p2, p3, ncol = 3)
```

### 2022-10-04 paper_comparison/paper_comparison/paper_comparison.tex
merged conflicts

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/residual-plot-common-departures-1} \caption{Example fitted vs residual plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted plot.}\label{fig:residual-plot-common-departures}
\end{figure}

@@

Example residual plots given in Figure \ref{fig:commondepartures} are
examined by the corresponding model structural test, Breusch-Pagan test
and Shapiro--Wilk test as shown in Table
\ref{tab:example-residual-plot-table}. In the example, both the
Breusch-Pagan test and the Shapiro--Wilk test rejects the null
hypothesis for departures that they do not intend to examine. As
discussed in \citet{cook1982residuals}, most residual-based tests for a
particular type of departure from model assumptions are sensitive to
other types of departures. It is likely the null hypothesis is correctly
rejected but for the wrong reason, which is known as the ``Type III
error''. Additionally, outliers will often incorrectly trigger the
rejection of the null hypothesis despite when majority of the residuals
are well-behaved \citep{cook_applied_1999}. This can be largely avoided
in diagnostic plots as experienced analysts can evaluate the
acceptability of assumptions flexibly, even in the presence of outliers.

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/first-example-lineup-1} \caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot 11, exhibiting non-linearity) is embedded among 19 null plots, where the residuals were simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot 11.}\label{fig:first-example-lineup}
\end{figure}

@@

\hypertarget{experiment-overview}{%
\subsection{Experiment overview}\label{experiment-overview}}

@@

Experiment I is designed to study the ability of human subjects to
detect the effect of a non-linear term \(\boldsymbol{z}\) which is a
probabilist's Hermite polynomial {[}Herimite ref here{]} of another
random vector \(\boldsymbol{x}\) in a two variable statistical model
formulated as:

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-shape-of-herimite-1} \caption{Example residual plots of the linear regression model used in experiment I. Four different values of $j$ are included in the experiment results in four different shapes of residuals.}\label{fig:different-shape-of-herimite}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-dist-1} \caption{Example residual plots of the linear regression model used in experiment I. Four different distribution of $x_{raw}$ are used in the experiment to provide various visual patterns.}\label{fig:different-dist}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/example-poly-lineup-1} \caption{Lineup poly-24 in experiment I. Can you spot the most different plot? \label{fig:example-poly-lineup}}\label{fig:example-poly-lineup}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-shape-of-heter-1} \caption{Example residual plots of the linear regression model used in experiment II. Three different shapes (a = -1, 0, 1) are used in the experiment to create left-triangle shape, butterfly shape and right-triangle shape.}\label{fig:different-shape-of-heter}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/example-heter-lineup-1} \caption{Lineup heter-169 in experiment II. Can you spot the most different plot? \label{fig:example-heter-lineup}}\label{fig:example-heter-lineup}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-n-1} \caption{Three different values of $n$ are used in experiment I, II and III to control the strength of the signal.}\label{fig:different-n}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-sigma-1} \caption{Example residual plots of the linear regression model used in experiment I. Four different values of $\sigma$ are included in the experiment to control the strength of the signal.}\label{fig:different-sigma}
\end{figure}

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-b-1} \caption{Example residual plots of the linear regression model used in experiment II. Five different values of $b$ are included in the experiment to control the strength of the signal.}\label{fig:different-b}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/fig:power-poly-uniform-1} \caption{...}\label{fig:fig:power-poly-uniform}
\end{figure}

@@

\begin{figure}
\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/dist-power-1} \caption{...}\label{fig:dist-power}
\end{figure}

\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/unnamed-chunk-7-1}

>>>>

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/residual-plot-common-departures-1}

}

\caption{Example fitted vs residual plots: (A) classically good looking residuals, (B) non-linear pattern indicates that the model has not captured a non-linear association, (C) heteroskedasticity indicating that variance around the fitted model is not uniform, and (D) non-normality where the residual distribution is not symmetric around 0. The latter pattern might best be assessed using a univariate plot of the residuals, but patterns B and C need to be assessed using a residual vs fitted plot.}\label{fig:residual-plot-common-departures}
\end{figure}

@@

Example residual plots given in Figure
\ref{fig:residual-plot-common-departures} are examined by the
corresponding model structural test, Breusch-Pagan test and
Shapiro--Wilk test as shown in Table
\ref{tab:example-residual-plot-table}. In the example, both the
Breusch-Pagan test and the Shapiro--Wilk test rejects the null
hypothesis for departures that they do not intend to examine. As
discussed in \citet{cook1982residuals}, most residual-based tests for a
particular type of departure from model assumptions are sensitive to
other types of departures. It is likely the null hypothesis is correctly
rejected but for the wrong reason, which is known as the ``Type III
error''. Additionally, outliers will often incorrectly trigger the
rejection of the null hypothesis despite when majority of the residuals
are well-behaved \citep{cook_applied_1999}. This can be largely avoided
in diagnostic plots as experienced analysts can evaluate the
acceptability of assumptions flexibly, even in the presence of outliers.

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/first-example-lineup-1}

}

\caption{Visual testing is conducted using a lineup, as in the example here. The residual plot computed from the observed data (plot 11, exhibiting non-linearity) is embedded among 19 null plots, where the residuals were simulated from a standard error model. Computing the $p$-value requires that the lineup be examined by a number of human judges, each asked to select the most different plot. A small $p$-value would result from a substantial number selecting plot 11.}\label{fig:first-example-lineup}
\end{figure}

@@

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-shape-of-herimite-1}

}

\caption{Polynomial forms generated for the residual plots used in experiment I. The four shapes are generated by varying the order of polynomial given by $j$ in $He_j(.)$.}\label{fig:different-shape-of-herimite}
\end{figure}

Experiment I is designed to study the ability of human subjects to
detect the effect of a non-linear term \(\boldsymbol{z}\) constructed
using Hermite polynomials {[}Hermite ref here{]} on random vector
\(\boldsymbol{x}\) formulated as:

@@

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-dist-1}

}

\caption{Variations in fitted values, that might affect perception of residual plots. Four different distribution of $x_{raw}$ are used in the experiment to provide various visual patterns.}\label{fig:different-dist}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/example-poly-lineup-1}

}

\caption{Lineup poly-24 in experiment I. Can you spot the most different plot? \label{fig:example-poly-lineup}}\label{fig:example-poly-lineup}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-shape-of-heter-1}

}

\caption{Heteroskedasticity forms used in experiment II. Three different shapes (a = -1, 0, 1) are used in the experiment to create left-triangle, butterfly and right-triangle shapes.}\label{fig:different-shape-of-heter}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/example-heter-lineup-1}

}

\caption{Lineup heter-169 in experiment II. Can you spot the most different plot? \label{fig:example-heter-lineup}}\label{fig:example-heter-lineup}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-n-1}

}

\caption{Three different values of $n$ are used in experiment I, II and III to control the strength of the signal.}\label{fig:different-n}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-sigma-1}

}

\caption{Example residual plots of the linear regression model used in experiment I. Four different values of $\sigma$ are included in the experiment to control the strength of the signal.}\label{fig:different-sigma}
\end{figure}

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/different-b-1}

}

\caption{Example residual plots of the linear regression model used in experiment II. Five different values of $b$ are included in the experiment to control the strength of the signal.}\label{fig:different-b}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/fig:power-poly-uniform-1}

}

\caption{...}\label{fig:fig:power-poly-uniform}
\end{figure}

@@

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/dist-power-1}

}

\caption{...}\label{fig:dist-power}
\end{figure}

\begin{center}\includegraphics[width=1\linewidth]{paper_comparison_files/figure-latex/unnamed-chunk-7-1} \end{center}
